---
title: "Private Health Insurance"
author: "Ian McCarthy | Emory University"
format: 
  revealjs:
    theme: [moon]
    preview-links: auto
    chalkboard:
      boardmarker-width: 5
#    embed-resources: true
from: markdown+emoji
execute: 
  echo: true
---

```{r}
#| include: false
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, lubridate, gapminder, scales)
```

# Table of contents

1. History of Health Insurance
2. Employer Sponsored Insurance
3. Non-group Coverage



# Background

---

## Early history of health insurance

- 1929: "Hospital insurance" for schoolteachers to receive care from Baylor University Hospital in Dallas, TX
- Essentially pre-paid medical care for about 1,500 schoolteachers in the area
- Annual fee (or "premium") of \$6 per year, \$104 in 2023, for up to 21 days of hospital care
- Subscription model offered regular cash flow for hospitals

---

- Popular revenue model for hospitals
- Network of hospitals grew to be Blue Cross, first established in Sacramento, CA in 1932
- Structure introduces a few distortions:
    - Care concentrated in hospitals, even if other settings could provide care at lower cost
    - No price competition
- Blue Cross plans grew in the 1930s
    - Operated as a not-for-profit
    - Exempt from typical regulations on insurance markets

---

- Expansion into prepaid physician services (Blue Shield), combined to be Blue Cross and Blue Shield
- By 1940, half of all health insurance plans were Blue Cross and Blue Shield plans
- Hospitals paid on a cost-plus basis (more on this later)
- Any competing insurance plans had to offer the same structure

---

## Modern health insurance

- Primarily through employers 
- Only developed country to rely on employer-sponsored insurance

```{r}
#| include: false
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, janitor, plotly)

ins.dat <- read_delim("../data/acs/acs_insurance.txt")

ins.dat <- ins.dat %>%
  mutate(perc_private = (ins_employer + ins_direct)/adult_pop,
         perc_public = (ins_medicare + ins_medicaid)/adult_pop,
         perc_ins = (adult_pop - uninsured)/adult_pop,
         perc_unins = uninsured/adult_pop,
         perc_employer = ins_employer/adult_pop,
         perc_medicaid = ins_medicaid/adult_pop,
         perc_medicare = ins_medicare/adult_pop,
         perc_direct = ins_direct/adult_pop)
```

```{r}
#| echo: false
#| fig.cap: "Source of Health Insurance (2012)"
#| label: fig-insurance-type
ins.dat %>%
  filter(year==2012) %>% pivot_longer(c("perc_employer","perc_direct","perc_medicaid", "perc_unins")) %>%
  select(State, name, value) %>%
  group_by(name) %>% summarize(mean=mean(value)) %>%
  ggplot(aes(x=name, y=mean)) +
  geom_col() + ylim(0,0.65) + 
  labs(
    x=" ",
    y="Percent",
    caption="Data from the American Community Survey, adults under age 65",
    title=" "
  ) + theme_bw() + scale_x_discrete(labels=c('perc_direct'='Direct Purchase',
                                             'perc_employer'='Employer-provided',
                                             'perc_medicaid'='Medicaid',
                                             'perc_unins'='Uninsured'))
```


---




